---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ATQ: Assesing Evaluation Metrics for Timely Epidemic Detection Models

## Background

Madeline A. Ward published a [paper](https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-019-7521-7) on methods for detecting seasonal influenza epidemics using school absenteeism data. The premise is that there is a school absenteeism surveillance system established in Wellington-Dufferin-Guelph which uses a threshold-based (10% of students absent) approach to raise alarms for school illness outbreaks to implement mitigating measures. Two metrics (FAR and ADD) are proposed in that study that were found to be more accurate.

Based on the work of Madeline, in 2021 Kayla Vanderkruk along with Drs. Deeth and Feng wrote a [paper](https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-023-15747-z) on improved metrics, namely ATQ (alert time quality), that are more accurate and timely than the FAR-selected models. This package is based off Kayla's work that can be found [here](https://github.com/vanderkk/School_Abstenteeism_Based_Influenza_Surveillance_Simulation_Study). ATQ study assessed alarms on a gradient, where alarms raised incrementally before or after an optimal date were penalized for lack of timeliness. 

This ATQ package will allow users to perform simulation studies and evaluate ATQ & FAR based metrics from them. The simulation study will require information from census data for a region such as distribution of number of household members, households with and without children, and age category, etc. 


This package is still a work in progress and future considerations include streamlining simulation study workflow and generalizing evaluation of metrics to include real data sets. 

## Installation:

You can install the development version of ATQ from [Github](https://github.com/vjoshy/ATQ_Surveillance_Package)

```r  
#install.packages("devtools")
library(devtools)
install_github("vjoshy/ATQ_Surveillance_Package")
```

## Key Functions

The ATQ package includes the following main functions:

1. `catchment_sim()`
    - Simulates catchment area data
2. `elementary_pop()`
    - Simulates elementary school populations
3. `subpop_children()`
    - Simulates households with children
4. `subpop_noChildren()`
    - Simulates households without children
5. `simulate_households()`
    - Combines household simulations
6. `ssir()`
    - Simulates epidemic using SSIR model
7. `compile_epi()`
    - Compiles absenteeism data
8. `eval_metrics()`
    - Evaluates alarm metrics
9. `plot()` and `summary()`
    - Methods for visualizing and summarizing results

Additionally, the package implements S3 methods for generic functions:

9. `plot()` and `summary()`
    - These methods are extended for objects returned by `ssir()` and `eval_metrics()`
    - `plot()` provides visualizations of epidemic simulations and metric evaluations
    - `summary()` offers concise summaries of the simulation results and metric assessments

These functions and methods work together to facilitate comprehensive epidemic simulation and evaluation of detection models


Please see example below:

```{r example}
# Load the ATQ package
library(ATQ)

#Simulate number of elementary schools in each catchment
catch_df <- catchment_sim(16, 4.68, 3.01, 5)

# Simulate elementary school populations for each catchment area
elementary_df <- elementary_pop(catch_df, 5.86, 0.01)

# Set up file connection to simulate user input for subpop_children() function
f <- file()
lines <- c(0.7668901,0.3634045, 0.4329440, 0.2036515,0.5857832, 0.3071523, 0.1070645,0.4976825)
ans <- paste(lines, collapse = "\n")
write(ans, f)
options("usr_con" = f) # Set connection option to use file instead of stdin

# Simulate households with children
house_children <- subpop_children(elementary_df, n = 2)

# Prepare input for subpop_noChildren() function
lines <- c(0.23246269, 0.34281716, 0.16091418, 0.16427239, 0.09953358, 0.4277052)
ans <- paste(lines, collapse = "\n")
write(ans, f)

# Simulate households without children
house_nochildren <- subpop_noChildren(house_children, elementary_df)

# Clean up file connection
close(f)
options("usr_con" = stdin()) # Reset connection option to standard input

# Combine household simulations and generate individual-level data
simulation <- simulate_households(house_children, house_nochildren)

# Extract individual-level data
individuals <- simulation$individual_sim

# Simulate epidemic using SSIR (Stochastic Susceptible-Infectious-Recovered) model
epidemic <- ssir(nrow(individuals), T = 300, alpha = 0.298, inf_init = 32, rep = 10)

# Summarize and plot the epidemic simulation results
summary(epidemic)
plot(epidemic)

# Compile absenteeism data based on epidemic simulation and individual data
absent_data <- compile_epi(epidemic, individuals)

# Display structure of absenteeism data
dplyr::glimpse(absent_data)

# Evaluate alarm metrics for epidemic detection
alarm_metrics <- eval_metrics(absent_data, thres = seq(0.1,0.6,by = 0.05),
                              ScYr = c(2:10), yr.weights = c(1:9)/sum(c(1:9)))

# Plot various alarm metrics
plot(alarm_metrics$metrics, "FAR")   # False Alert Rate
plot(alarm_metrics$metrics, "FATQ")  # First Alert Time Quality
plot(alarm_metrics$metrics, "AATQ")  # Average Alert Time Quality
plot(alarm_metrics$metrics, "WAATQ") # Weighted Average Alert Time Quality
plot(alarm_metrics$metrics, "WFATQ") # Weighted First Alert Time Quality
plot(alarm_metrics$metrics, "ADD")   # Accumulated Delay Days


# Summarize alarm metrics
summary(alarm_metrics$summary)

# Generate and display plots for alarm metrics across epidemic years
alarm_plots <- plot(alarm_metrics$plot_data)
for(i in seq_along(alarm_plots)) { 
  print(alarm_plots[[i]]) 
}
```

The final output 'region_metric' will be a list of 6 matrices and 6 data frames. The matrices describe the values of metrics for respective lag and thresholds. 

An optimal lag and threshold value that minimizes each metric is selected and these optimal parameters are used to generate the 6 data frames associated with the metrics. Simulated information like number of lab confirmed cases, number of students absent, etc, are included in the output.

These data frames contain 19 variables and are as follows:

  *  lag, Best selected lagged absenteeism value.
  *  thres, Best selected threshold value.
  *  Data, Day (1-300).
  *  ScYr, School year.
  *  Actual.case, Total number of true cases (unconfirmed).
  *  case.elem, Number of true cases that are elementary school children.
  *  Case.No, Number of lab confirmed cases.
  *  Case, Binary variable to indicate whether there has been a lab confirmed case for the day.
  *  pct.absent, Percent of students absent in total school population.
  *  absent, Total number of students absent.
  *  absent.sick, Total number of students absent due to illness.
  *  window, 14 day true alarm window for ADD and FAR calculations (binary variable).
  *  ref.date, Reference date (first day where there are two lab confirmed cases in a 7 day period).
  *  Alarm, Binary variable indicating whether alarm was raised (1 - raised, 0 - not raised).
  *  ATQ, Alert time quality (metric).
  *  FAR, False alarm rate (metric).
  *  ADD, Accumulated days delayed (metric).
  *  AATQ, Average alarm time quality (metric).
  *  FATQ, First alarm time quality (metric).
